---
title: "Core Covid-19 Stats for Denmark"
date: 2022-05-24
output:
  html_document:
    toc: true
---
```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('core.R')
```

# Covid 19 Stats for Denmark

This file provides the new development of Covid-19 stats. Where the original development took its data from the SSI Covid-19 Dashboard, this takes data in the form of the "raw" core data.

One early important part of the process is to define a date range of interest. The current variant we are interested in is Omicron, so, we can focus on recent data, rather than the full period:

```{r cutoffs}
```

Likewise, some files contain a date at which they were produced. For instance the estimation of R~t~:

```{r dates}
```

## How data is read and handled

The R site we are creating here uses a separate devleopment, `core.R` to read in and handle core data. This document then references into the core R development when needed for different code chunks. This allows us for easy scripting and exploration in the R session while also neatly handling dissemination in a document like this.

We first need a couple of libraries

```{r libraries, message=FALSE}
```

Where the Dashboard data is stored in ISO8859-1 format, the core data is already in UTF-8. So they are way easier to read. We can just point tidyverse's `readr` package directly to the data at hand. Start by setting up a core path

```{r where_we_are}
```

### Case Estimation of R~t~

Read in the R~t~ case estimation data

```{r read_rt, message=FALSE}
```

### Municipality cases and tests as a time series

This data is easily read:

```{r read_m_cases_timeseries, message = FALSE}
```

However, as we see, the data is *messy* and not *tidy*. To make the data set tidy, we want to bring it on essentially 3rd normal form:

```{r cleanup_m_cases_timeseries}
```

We use the same strategy for the tests:

```{r read_m_tested_timeseries, message = FALSE}
```

Since these two tibbles are storing the same data, we can join them:

```{r join_m_timeseries}
```

## Plots of interest

### Cases in different municipalities

Generally, the most recent observations in the data sets are flawed. Their counts tend to increase over time as data is reported into the central register. Hence, it is wise to cut these data from the data set and look at the remaining data instead.

```{r m_high_cutoff}
```

Furthermore, the we have few observations from island-data. Because of this, it is wise to filter them out, so we can get a better picture of the remaining municipalities. Split off these data sets into their own:

```{r m_split_off_islands}
```

Plot for the "mainland"

```{r m_plot_rate, fig.height=12}
```

Plot for the islands

```{r m_plot_rate_islands}
```

Country-wide tests, cases, positive-rate
```{r m_summarize}
```

### Regions

Municipalities are grouped by region, which are defined by the following vectors:
```{r regional_area_vectors}
```

Once we have the vectors, we can construct a tibble from them:

```{r region_mapping}
```

And joining that tibble to our data, we can plot per region
```{r region_split, fig.height=12, fig.width=16}
```

Finally, we can plot positive rate for each region
```{r region_split_rate, fig.height=12, fig.width=16}
```
