---
title: "Core Covid-19 Stats for Denmark"
date: 2022-01-31
output:
  html_document:
    toc: true
---
```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('core.R')
```

# Covid 19 Stats for Denmark

This file provides the new development of Covid-19 stats. Where the original development took its data from the SSI Covid-19 Dashboard, this takes data in the form of the "raw" core data.

One early important part of the process is to define a date range of interest. The current variant we are interested in is Omicron, so, we can focus on recent data, rather than the full period:

```{r cutoffs}
```

## How data is read and handled

The R site we are creating here uses a separate devleopment, `core.R` to read in and handle core data. This document then references into the core R development when needed for different code chunks. This allows us for easy scripting and exploration in the R session while also neatly handling dissemination in a document like this.

We first need a couple of libraries

```{r libraries, message=FALSE, echo = FALSE}
```

Where the Dashboard data is stored in ISO8859-1 format, the core data is already in UTF-8. So they are way easier to read. We can just point tidyverse's `readr` package directly to the data at hand. Start by setting up a core path

```{r where_we_are}
```

### Municipality cases as a time series

This data is easily read:

```{r read_m_cases_timeseries}
```

However, as we see, the data is *messy* and not *tidy*. To make the data set tidy, we want to bring it on essentially 3rd normal form:

```{r cleanup_m_cases_timeseries}
```

## Plots of interest

### Cases in different municipalities

```{r m_plot_cases, fig.height=12}
```
