---
title: "Covid-19 Stats for Denmark"
date: 2022-01-13
output:
  html_document:
    toc: true
---

## Initial setup

To process data sets from SSI, we'll use R, and we'll need some packages for doing so. We'll
load `tidyverse` to load a large set of R packages, and `lubridate` to load in date-handling,
since this will be very useful.

You can grab SSI's data from [their homepage](https://covid19.ssi.dk/overvagningsdata/download-fil-med-overvaagningdata)

```{r setup, include = TRUE, echo = TRUE, message = FALSE}
library(tidyverse)
library(lubridate)
```

SSI's data is in a zip file. We unpack this zip file into a directory of the
general form `YYYY-MM-DD`. Lets set up a name for the current directory we
want to grab data from

```{r constants}
date <- "2022-01-12"
```

## Ingesting data

SSIs data are in "CSV" files. The CSV format isn't well specified, and the variant
used by SSI is a variant in which the `;` character is used as a separator. This is
relatively common in countries which uses the `,` as a separator in numbers, in order
to avoid conflicts. R's `readr` package have a way to handle this particular CSV variant.

Furthermore, the files contain danish characters. They are encoded in the ISO8859-1(5)
format, so we'll have to do something about this. The obvious way is to turn the
encoding into UTF-8, which is the modern encoding of Unicode. Once transformed into
UTF-8, we can read the data into R.

We'll capture this conversion into a function `read_data`. It uses
[iconv](https://en.wikipedia.org/wiki/Iconv) to convert between encodings. The function
also allows us to designate the types of the columns in the file we are trying to read.

```{r read_function}
read_data <- function(f, col_types) {
    raw <- readBin(f, "raw", n = file.size(f))
    x_utf8 <- iconv(list(raw), from = "ISO-8859-1", to = "UTF-8", toRaw = TRUE)[[1]]

    # CSV variant 2 uses ; as the separator, among other adaptations
    res <- read_csv2(x_utf8, col_types=col_types)
    return(res)
}
```

### Key Numbers

SSI provides a file with key numbers from the Covid-19 data gathered by them. This file is updated 5
days a week. We already unzipped the file in a directory, so lets expand our `date` constant and get a
file name

```{r}
file <- sprintf("./data/%s/Regionalt_DB/03_bekraeftede_tilfaelde_doede_indlagte_pr_dag_pr_koen.csv", date)
```

We can read this file via our function. It will return a so-called `tibble` which is a table, or data frame,
of the data in the file. We also provide initial types for each of the columns. In CSV, data is untyped.
We would like to add type information to columns because it simplifies our later plotting. Many tools
know how to handle data by its type, so we designate things up front.

```{r}
data <- read_data(file, col_types=cols('Region' = col_factor(),
                                       'Prøvetagningsdato' = col_date(),
                                       'Køn' = col_factor(),
                                       'Bekræftede tilfælde' = col_double(),
                                       'Døde' = col_double(),
                                       'Indlæggelser' = col_double()
                                      ))
data
```

# Danish in a file is a miserable situation. Convert into the internets lingua franca,
# i.e., english:
data <- data %>% rename(region = Region,
                        date = Prøvetagningsdato,
                        gender = Køn, # This might be wrong, because it could also be 'sex'. However, we are going to aggregate it
                        cases = 'Bekræftede tilfælde',
                        deaths = Døde,
                        hospitalizations = Indlæggelser)

# Cut the cumulative columns away
data <- data %>% select(region, date, gender, cases, deaths, hospitalizations)

# We are only interested in somewhat recent data. What happened in the past isn't that
# important
lo_cutoff <- '2021-09-01'
hi_cutoff <- max(data$date) - days(2)
data <- data %>% filter(date >= cutoff, date < hi_cutoff)

# Put this into the summary
summary <- data

file <- sprintf("./data/%s/Regionalt_DB/18_fnkt_alder_uge_testede_positive_nyindlagte.csv", date)
# Parse data, by reading the CSV file, and do column type conversion
# This makes a lot of later handling easier because we have actual types
# for the data in our tibble
data <- read_data(file, col_types=cols(Uge = col_character(), # This will need fixing later
                                       Aldersgruppe = col_factor(),
                                       'Testede pr. 100.000 borgere' = col_double(),
                                       'Positive pr. 100.000 borgere' = col_double(),
                                       'Nyindlagte pr. 100.000 borgere' = col_double(),
                                       'Antal testede' = col_double(),
                                       'Antal positive' = col_double()
                                      ))

# Work on the data, and start fixing. Step 1 is to clean up the column names
data <- data %>% rename(week = Uge,
                        ageGroup = Aldersgruppe,
                        tested_100k = 'Testede pr. 100.000 borgere',
                        cases_100k = 'Positive pr. 100.000 borgere',
                        hospitalized_100k = 'Nyindlagte pr. 100.000 borgere',
                        tested = 'Antal testede',
                        cases = 'Antal positive')

# Next step is to handle the week column. Define a fixup function:
fixup_week <- function(s) {
    # For strptime to do its work correctly, it must know a day of the week. Here, we
    # use the 1st day of the week as a anchoring point, but in principle any day could
    # have been used.
    x <- sprintf("%s-1", s)
    res <- strptime(x, format="%Y-W%W-%u")
    return(as.POSIXct(res))
}

# Apply this to the data set
data <- data %>% mutate(week = fixup_week(week))
data <- data %>% filter(week >= cutoff)

# Shove this into a new variable for our use
weekly <- data

### COUNTRY-WIDE data

# Group by date, so we can get rid of region/gender
country <- summary %>% group_by(date)
# Summarize the data by counting
country <- country %>% summarize(cases = sum(cases),
                                 deaths = sum(deaths),
                                 hospitalizations = sum(hospitalizations))

# Create a lag column for cases
country <- country %>% mutate(previous_cases = lag(cases, order_by=date),
                              weekday = lubridate::wday(date, label=TRUE))

plot_country <- function(p, filename, title, subtitle) {
    p + geom_point(aes(color = weekday), size=3) +
        geom_smooth(method = 'loess', span = 0.3) +
        scale_colour_brewer(palette="Set2") +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

# Write to a file
p <- ggplot(country, aes(x=date, y=cases))
plot_country(p, "cases.png", "Covid-19 cases in Denmark", "Source: ssi.dk")

# Write to a file
p <- ggplot(country, aes(x=date, y=hospitalizations))
plot_country(p, "hospitalizations.png", "Covid-19: new hospitalizations in Denmark", "Source: ssi.dk")

p <- ggplot(country, aes(x=date, y=deaths))
plot_country(p, "deaths.png", "Covid-19: new deaths in Denmark", "Source: ssi.dk")

# Plot per weekday
plot_day_of_week <- function(p, filename, title, subtitle) {
    p + geom_point() +
        geom_jitter() +
        scale_colour_brewer(palette="Set1") +
        stat_boxplot(aes(fill=weekday, color=weekday), alpha=0.3) +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

p <- ggplot(country, aes(x=weekday, y=(cases - previous_cases)))
plot_day_of_week(p, "cases_weekday.png", "New Covid-19 cases in Denmark, day of week (2021-09-01 to 2022-01-10)", "Source: ssi.dk")

# Lagged data computes the difference to the day before.
plot_country_lag <- function(p, filename, title, subtitle) {
    p + geom_point(size=3) +
        scale_colour_brewer(palette="Set2") +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

p <- ggplot(country, aes(x=date, y=(cases - previous_cases), color=weekday))
plot_country_lag(p, "cases_lag.png", "Covid-19 cases change per day, Denmark", "Source: ssi.dk")

### REGIONAL DATA
## Generally follows the same structure as for the country-wide cases.

regions <- summary %>% group_by(date, region)
regions <- regions %>% summarize(cases = sum(cases), deaths = sum(deaths), hospitalizations = sum(hospitalizations))
regions <- regions %>% mutate(weekday = lubridate::wday(date, label=TRUE), order_by=date)

plot_regions <- function(p, filename, title, subtitle) {
    p + geom_point(aes(color=region), size=2) +
        geom_smooth(method = 'loess', span = 0.3) +
        facet_grid(rows=vars(region)) +
        scale_colour_brewer(palette="Set2") +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

p <- ggplot(regions, aes(x=date, y=cases))
plot_regions(p, "regions.png", "Covid-19 cases in Denmark, per region", "Source: ssi.dk")

plot_age_cases <- function(p, filename, title, subtitle) {
    p + geom_raster() +
        scale_fill_continuous(type = "viridis") +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

p <- ggplot(weekly, aes(x=week, y=ageGroup, fill=cases_100k))
plot_age_cases(p, "weekly_age_cases.png", "Covid-19 cases per 100k in Denmark, by age group", "Source: ssi.dk")

p <- ggplot(weekly, aes(x=week, y=ageGroup, fill=tested_100k))
plot_age_cases(p, "weekly_age_tested.png", "Covid-19 tests per 100k in Denmark, by age group", "Source: ssi.dk")
