---
title: "Covid-19 Stats for Denmark"
date: 2022-01-13
output:
  html_document:
    toc: true
---

## Initial setup

To process data sets from SSI, we'll use R, and we'll need some packages for doing so. We'll
load `tidyverse` to load a large set of R packages, and `lubridate` to load in date-handling,
since this will be very useful.

You can grab SSI's data from [their homepage](https://covid19.ssi.dk/overvagningsdata/download-fil-med-overvaagningdata)

```{r setup, include = TRUE, echo = TRUE, message = FALSE}
library(tidyverse)
library(lubridate)
```

SSI's data is in a zip file. We unpack this zip file into a directory of the
general form `YYYY-MM-DD`. Lets set up a name for the current directory we
want to grab data from

```{r constants}
date <- "2022-01-12"
```

## Ingesting data

SSIs data are in "CSV" files. The CSV format isn't well specified, and the variant
used by SSI is a variant in which the `;` character is used as a separator. This is
relatively common in countries which uses the `,` as a separator in numbers, in order
to avoid conflicts. R's `readr` package have a way to handle this particular CSV variant.

Furthermore, the files contain danish characters. They are encoded in the ISO8859-1(5)
format, so we'll have to do something about this. The obvious way is to turn the
encoding into UTF-8, which is the modern encoding of Unicode. Once transformed into
UTF-8, we can read the data into R.

We'll capture this conversion into a function `read_data`. It uses
[iconv](https://en.wikipedia.org/wiki/Iconv) to convert between encodings. The function
also allows us to designate the types of the columns in the file we are trying to read.

```{r read_function}
read_data <- function(f, col_types) {
    raw <- readBin(f, "raw", n = file.size(f))
    x_utf8 <- iconv(list(raw), from = "ISO-8859-1", to = "UTF-8", toRaw = TRUE)[[1]]

    # CSV variant 2 uses ; as the separator, among other adaptations
    res <- read_csv2(x_utf8, col_types=col_types)
    return(res)
}
```

### Key Numbers

SSI provides a file with key numbers from the Covid-19 data gathered by them. This file is updated 5
days a week. We already unzipped the file in a directory, so lets expand our `date` constant and get a
file name

```{r}
file <- sprintf("./data/%s/Regionalt_DB/03_bekraeftede_tilfaelde_doede_indlagte_pr_dag_pr_koen.csv", date)
```

We can read this file via our function. It will return a so-called `tibble` which is a table, or data frame,
of the data in the file. We also provide initial types for each of the columns. In CSV, data is untyped.
We would like to add type information to columns because it simplifies our later plotting. Many tools
know how to handle data by its type, so we designate things up front.

```{r}
data <- read_data(file, col_types=cols('Region' = col_factor(),
                                       'Prøvetagningsdato' = col_date(),
                                       'Køn' = col_factor(),
                                       'Bekræftede tilfælde' = col_double(),
                                       'Døde' = col_double(),
                                       'Indlæggelser' = col_double()
                                      ))
data
```

We certainly don't want to work with danish-named columns. Rather, we want small
descriptive names which are easy to type. We send our dataset through a `rename`
pass for this.

And while we are here, lets also limit ourselves to the columns we are interested
in. There are some cumulative columns in the dataset, but we are not going to be
interested in those. So we select the columns we care about.

```{r key_rename}
data <- data %>% rename(region = 'Region',
                        date = 'Prøvetagningsdato',
                        gender = 'Køn', # This might be wrong, because it could also be 'sex'. However, we are going to aggregate it
                        cases = 'Bekræftede tilfælde',
                        deaths = 'Døde',
                        hospitalizations = 'Indlæggelser')

# Cut the cumulative columns away
data <- data %>% select(region, date, gender, cases, deaths, hospitalizations)
data
```

More data handling. We are interested in somewhat recent data. Covid-19 has been around
for a while. We are mostly interested in the recent development which is that Delta was dominant,
and Omicron came into the space. Let's put a low-end cutoff down:

```{r}
lo_cutoff <- '2021-09-01'
```

The hi-end of that cutoff, we set to two days before the maximal date in our data set

```{r}
hi_cutoff <- max(data$date) - days(2)
```

Rationale: when you have a reported number in the news, it's generally a number from the
cumulative count. The actual date of testing is often lagging by 24h to 48h. We are interested
in getting the right test-date, even if it means we have a 2 days lag. Hence, we assume
that recent data are still "forming". A typical example is that the count for the current
date is very close to 0, as no tests have come in yet. In 2-3 days time, this number rises
to the right value, as data is retroactively updated.

Filtering the data set is easy:

```{r summary}
data <- data %>% filter(date >= lo_cutoff, date < hi_cutoff)

# Put this into the summary
summary <- data
summary
```

### Weekly key data split by age

As in the case with the general data, we can define the file of interest

```{r}
file <- sprintf("./data/%s/Regionalt_DB/18_fnkt_alder_uge_testede_positive_nyindlagte.csv", date)
```

We can reuse our data reader from before.

```{r}
data <- read_data(file, col_types=cols(Uge = col_character(), # This will need fixing later
                                       Aldersgruppe = col_factor(),
                                       'Testede pr. 100.000 borgere' = col_double(),
                                       'Positive pr. 100.000 borgere' = col_double(),
                                       'Nyindlagte pr. 100.000 borgere' = col_double(),
                                       'Antal testede' = col_double(),
                                       'Antal positive' = col_double()
                                      ))
```

# Work on the data, and start fixing. Step 1 is to clean up the column names
data <- data %>% rename(week = Uge,
                        ageGroup = Aldersgruppe,
                        tested_100k = 'Testede pr. 100.000 borgere',
                        cases_100k = 'Positive pr. 100.000 borgere',
                        hospitalized_100k = 'Nyindlagte pr. 100.000 borgere',
                        tested = 'Antal testede',
                        cases = 'Antal positive')

# Next step is to handle the week column. Define a fixup function:
fixup_week <- function(s) {
    # For strptime to do its work correctly, it must know a day of the week. Here, we
    # use the 1st day of the week as a anchoring point, but in principle any day could
    # have been used.
    x <- sprintf("%s-1", s)
    res <- strptime(x, format="%Y-W%W-%u")
    return(as.POSIXct(res))
}

# Apply this to the data set
data <- data %>% mutate(week = fixup_week(week))
data <- data %>% filter(week >= cutoff)

# Shove this into a new variable for our use
weekly <- data

### COUNTRY-WIDE data

# Group by date, so we can get rid of region/gender
country <- summary %>% group_by(date)
# Summarize the data by counting
country <- country %>% summarize(cases = sum(cases),
                                 deaths = sum(deaths),
                                 hospitalizations = sum(hospitalizations))

# Create a lag column for cases
country <- country %>% mutate(previous_cases = lag(cases, order_by=date),
                              weekday = lubridate::wday(date, label=TRUE))

plot_country <- function(p, filename, title, subtitle) {
    p + geom_point(aes(color = weekday), size=3) +
        geom_smooth(method = 'loess', span = 0.3) +
        scale_colour_brewer(palette="Set2") +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

# Write to a file
p <- ggplot(country, aes(x=date, y=cases))
plot_country(p, "cases.png", "Covid-19 cases in Denmark", "Source: ssi.dk")

# Write to a file
p <- ggplot(country, aes(x=date, y=hospitalizations))
plot_country(p, "hospitalizations.png", "Covid-19: new hospitalizations in Denmark", "Source: ssi.dk")

p <- ggplot(country, aes(x=date, y=deaths))
plot_country(p, "deaths.png", "Covid-19: new deaths in Denmark", "Source: ssi.dk")

# Plot per weekday
plot_day_of_week <- function(p, filename, title, subtitle) {
    p + geom_point() +
        geom_jitter() +
        scale_colour_brewer(palette="Set1") +
        stat_boxplot(aes(fill=weekday, color=weekday), alpha=0.3) +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

p <- ggplot(country, aes(x=weekday, y=(cases - previous_cases)))
plot_day_of_week(p, "cases_weekday.png", "New Covid-19 cases in Denmark, day of week (2021-09-01 to 2022-01-10)", "Source: ssi.dk")

# Lagged data computes the difference to the day before.
plot_country_lag <- function(p, filename, title, subtitle) {
    p + geom_point(size=3) +
        scale_colour_brewer(palette="Set2") +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

p <- ggplot(country, aes(x=date, y=(cases - previous_cases), color=weekday))
plot_country_lag(p, "cases_lag.png", "Covid-19 cases change per day, Denmark", "Source: ssi.dk")

### REGIONAL DATA
## Generally follows the same structure as for the country-wide cases.

regions <- summary %>% group_by(date, region)
regions <- regions %>% summarize(cases = sum(cases), deaths = sum(deaths), hospitalizations = sum(hospitalizations))
regions <- regions %>% mutate(weekday = lubridate::wday(date, label=TRUE), order_by=date)

plot_regions <- function(p, filename, title, subtitle) {
    p + geom_point(aes(color=region), size=2) +
        geom_smooth(method = 'loess', span = 0.3) +
        facet_grid(rows=vars(region)) +
        scale_colour_brewer(palette="Set2") +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

p <- ggplot(regions, aes(x=date, y=cases))
plot_regions(p, "regions.png", "Covid-19 cases in Denmark, per region", "Source: ssi.dk")

plot_age_cases <- function(p, filename, title, subtitle) {
    p + geom_raster() +
        scale_fill_continuous(type = "viridis") +
        labs(title = title, subtitle = subtitle)

    ggsave(filename, height=9, width=16, dpi=120)
}

p <- ggplot(weekly, aes(x=week, y=ageGroup, fill=cases_100k))
plot_age_cases(p, "weekly_age_cases.png", "Covid-19 cases per 100k in Denmark, by age group", "Source: ssi.dk")

p <- ggplot(weekly, aes(x=week, y=ageGroup, fill=tested_100k))
plot_age_cases(p, "weekly_age_tested.png", "Covid-19 tests per 100k in Denmark, by age group", "Source: ssi.dk")
